Deployable Probabilistic Programming

### 2019-04-20 paper/paper.tex
highlights in progress

Infergo employs reverse-mode autodifferentiation via source code
transformation~\cite{GW08}. The reverse mode is the default
choice in machine learning applications, because it works best
with differentiating a scalar return value with respect to
multiple parameters. Automatic differentiation is usually
implemented using either operator overloading or source code
transformation. Operator overloading is often the method of
choice~\cite{Stan17,GXG18,PGC+17} due to relative simplicity,
but the differentiated code must use a special type (on which
the operators are overloaded) for representing floating point
numbers. Regardless of the issue of computational overhead, this
restricts the language used to specify the model to operations on
that special type and impairs interoperability: library
functions cannot be directly called but have to use an adaptor,
or a `lifted' version of the standard library must be supplied.
Similarly, the data must be cast into the special type before it
is passed to the model for inference.

Source code transformation allows to automatically differentiate
program code operating on the built-in floating point type. This
method involves generation of new code, for the function
gradient or for the function itself, ahead of time or on the
fly. To achieve the best performance, some implementations
generate static code for computing the function
gradient~\cite{MMW18,I18}. This requires a rather elaborated
source code analysis, and although a significant progress has
been made in this direction, not all language constructs are
supported. Alternatively, the code of the function itself can be
transformed by augmenting arithmetics and function calls on the
floating point type by function calls that record the
computations on the \textit{tape}, just like in the case of
operator overloading. Then, the backward pass is identical to
that of operator overloading. This method allows the
use of the built-in floating point type along with
straightforward virtually unrestricted language for model
specification, at the cost of slight performance loss.

This last method is used in Infergo: the model is written in Go
and is placed in a separate Go package. A command-line utility
is run on the package to generate a sub-package containing a
differentiated version of the model.  Both the original and the
differentiated model are available simultaneously in the calling
code, so that invocations of model methods which do not require
computation of gradients can be performed without the overhead
of automatic differentiation.  However, the performance loss is
rarely a major issue --- differentiated versions of model
methods are only 2--4 times slower on models from the case
studies~\ref{sec:case-studies}, and the backward pass, which
computes the gradient, is roughly as fast as the original,
undifferentiated version. At least partially, fast backward pass
is due to careful implementation and thorough profiling of
the tape.

>>>>

Infergo employs reverse-mode automatic differentiation via
source code transformation~\cite{GW08}. The reverse mode is the
default choice in machine learning applications, because it
works best with differentiating a scalar return value with
respect to multiple parameters. Automatic differentiation is
usually implemented using either operator overloading or source
code transformation. Operator overloading is often the method of
choice~\cite{Stan17,GXG18,PGC+17} due to relative simplicity,
but the differentiated code must use a special type (on which
the operators are overloaded) for representing floating point
numbers. Regardless of the issue of computational overhead, this
restricts the language used to specify the model to operations
on that special type and impairs interoperability: library
functions cannot be directly called but have to use an adaptor,
or a `lifted' version of the standard library must be supplied.
Similarly, the data must be cast into the special type before it
is passed to the model for inference.

\paragraph{Reverse mode source code transformation} Source code
transformation allows to automatically differentiate program
code operating on the built-in floating point type. This method
involves generation of new code, for the function gradient or
for the function itself, ahead of time or on the fly. To achieve
the best performance, some implementations generate static code
for computing the function gradient~\cite{MMW18,I18}. This
requires a rather elaborated source code analysis, and although
a significant progress has been made in this direction, not all
language constructs are supported. Alternatively, the code of
the function itself can be transformed by augmenting arithmetics
and function calls on the floating point type by function calls
that record the computations on the \textit{tape}, just like in
the case of operator overloading. Then, the backward pass is
identical to that of operator overloading. This method allows
the use of the built-in floating point type along with
straightforward virtually unrestricted language for model
specification, at the cost of slight performance loss.

This last method is used in Infergo: the model is written in Go
and is placed in a separate Go package. A command-line utility
is run on the package to generate a sub-package containing a
differentiated version of the model.  Both the original and the
differentiated model are available simultaneously in the calling
code, so that model methods which do not require computation of
gradients can be invoked without the overhead of automatic
differentiation.  However, the performance loss is rarely a
major issue --- differentiated versions of model methods are
only 2--4 times slower on models from the case studies
(Section~\ref{sec:case-studies}), and the backward pass, which
computes the gradient, is roughly as fast as the original,
undifferentiated version. At least partially, fast backward pass
is due to careful implementation and thorough profiling of the
tape.

\paragraph{Automatic differentiation of models}
It may seem desirable to implement or use an automatic
differentiation library that is more general than needed for
probabilistic programming. However the use of such library, in
particular, with source code transformation, may make
implementation restricted and less efficient: first, if a model
comprises several functions, all functions must have a signature
suitable for differentiation (that is, accept and return
floating point scalars or vectors); second, the gradient of
every nested call must be fully computed first and then used to
compute the gradient of the caller. In Infergo, only the
gradient of \lstinline{Observe} is needed, and any other
differentiated model method can only be called from
\lstinline{Observe} or another differentiated method.
Consequently, only the gradient of \lstinline{Observe} is
computed explicitly, and other differentiated methods may have
arbitrary arguments. In particular, derivatives may propagate
through fields of the model object and structured arguments.

\subsection{Model composition}

Just like it is possible to compose library and user-defined
functions into more complex functions, it should be possible to
compose models to build other models~\cite{THS+17,SSW+18}.
Infergo supports model composition organically: a differentiated
model method can be called from another differentiated method,
not necessarily a method of the same model object or type.
Derivatives can propagate through methods of different models
without restriction. Each model is a building block for another
model. For example, if there are two independent
single-parameter models \lstinline{A} and \lstinline{B}
describing the data, then the composite model \lstinline{AB} can
be obtained by summation:
\begin{lstlisting}
func (model AB) Observe(x []float64) float64 {
	return A{model.Data}.Observe(x[:1]) + 
		B{model.Data}.Observe(x[1:])
}
\end{lstlisting}
Combinators such as product, mixture, or hierarchy can be
implemented as functions operating on Infergo models.
